Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-19 main.Rmd
data structure section update

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. However, restructuring time in this manner leads to re-organisation of the data structure, where each level of a cyclic granularity correspond to multiple values of the measured variable, unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see consolidation of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. But that approach  hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the probability distribution of these multiple observations across cyclic granularities using some summary statistics together, each of which can highlight different features of the data could be a potential way to comprehend periocities in the data. Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time.

@@

## Data structure for exploring two cyclic granularities {#sec:twocylcic}

A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in (Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities. Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ denotes the measurement variable. Data sets of the form <$C_1$, $C_2$, $v$> can then form the basis for exploration and analysis of the measured variable across bivariate cyclic granularities.

@@

### Summarize distribution of measurements {#sec:summarize-measured}

Summarizing properties of the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. However, we should have the consider the effect of number of observations on the summarization even for harmonies. Suppose we have $T$ observations, and two cyclic granularities $C_1$ with $n$ categories and $C_2$ with $m$ categories. Each element of $C_1$ occurs approximately $T/n$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(mn)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge mnk$, the visualization will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, we would need $k \ge 30$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

>>>>

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in (Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities. Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ denotes the measurement variable. Data sets of the form <$C_1$, $C_2$, $v$> can then form the basis for exploration and analysis of the measured variable across bivariate cyclic granularities.

@@

@@

### Summarizing the measured variable {#sec:summarize-measured}

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity correspond to multiple values of the measured variable. This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. But that approach  hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of these multiple observations across cyclic granularities using some summary statistics, each of which can highlight different features of the data or through probability densities could be a potential way to compare and contrast periocities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

However, we need to consider the effect of number of observations on the summarization even for harmonies. Suppose we have $T$ observations, and two cyclic granularities $C_1$ with $n$ categories and $C_2$ with $m$ categories. Each element of $C_1$ occurs approximately $T/n$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(mn)$. If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge mnk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.
